Practical Hilbert space approximate Bayesian Gaussian processes for probabilistic programming

### 2022-02-09 Paper/Manuscript_Stats&Comp/manuscript_STCO.tex
update manuscript

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
%\begin{filecontents*}{example.eps}
%%!PS-Adobe-3.0 EPSF-3.0
%%%BoundingBox: 19 19 221 221
%%%CreationDate: Mon Sep 29 1997
%%%Creator: programmed by hand (JK)
%%%EndComments
%gsave
%newpath
%  20 20 moveto
%  20 220 lineto
%  220 220 lineto
%  220 20 lineto
%closepath
%2 setlinewidth
%gsave
%  .4 setgray fill
%grestore
%stroke
%grestore
%\end{filecontents*}
%
%\RequirePackage{fix-cm}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
%\documentclass[smallextended]{svjour3}       % onecolumn (second format)
\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
\usepackage{blindtext}
%\usepackage{titlesec}
\usepackage[utf8]{inputenc}
\usepackage{times}
\usepackage{microtype}
\usepackage{lscape}
\usepackage{epstopdf}
\usepackage{natbib}
\usepackage{subfigure}
\usepackage{multirow}
\usepackage{multicol}
\usepackage{float}
\usepackage{soul}
\usepackage{xcolor}
\usepackage{amssymb}
\usepackage{amsmath}	
\usepackage{mathtools}
%\usepackage{amsthm}
\usepackage{bm}
\usepackage{url}
\usepackage{parskip}
\usepackage{pbox}
\usepackage[titletoc]{appendix}
\usepackage{booktabs}
\usepackage{boldline}
\usepackage{colortbl}
\usepackage{listings}
\usepackage{caption}
\usepackage[colorlinks=true,citecolor=black,linkcolor=black,urlcolor=black,breaklinks=true]{hyperref}

@@

@@

The Mat\'ern class of isotropic covariance functions is given by, 
%
\begin{align*}
k_{\nu}(r)&=\alpha \, \frac{2^{1-\nu}}{\Gamma(\nu)}\left(\frac{\sqrt{2\nu}r}{\ell}\right)^{\!\nu} \! K_{\nu} \left(\frac{\sqrt{2\nu}r}{\ell}\right),
\end{align*}
%
where $\nu > 0$ is the order the kernel, $K_{\nu}$ the modified Bessel function of the second kind, and the $\ell > 0$ and $\alpha > 0$ are the length-scale and magnitude (marginal variance), respectively, of the kernel. The particular case where $\nu=\infty$ and $\nu=3/2$ are probably the most commonly used kernels \citep{rasmussen2006gaussian}, 
%
\begin{align*}
k_{\infty}(r)&=\alpha \exp \left( -\frac{1}{2} \frac{r^2}{\ell^2}\right),  \\
k_{\frac{3}{2}}(r)&=\alpha\left(1+\frac{\sqrt{3}r}{\ell}\right) \exp\! \left( -\frac{\sqrt{3}r}{\ell}\right). 
\end{align*}
%
The former is commonly known as the squared exponential or exponentiated quadratic covariance function. Assuming the Euclidean distance between observations, $r=|\bm{x}-\bm{x}'|_{L2}=\sqrt{\sum_{i=1}^{D}(x_i-x_i')^2}$, the kernels written above take the form
%
\begin{align*}
k_{\infty}(r) = \alpha & \exp \left( -\frac{1}{2} \sum_{i=1}^{D}\frac{(x_i-x_i')^2}{\ell_i^2}\right),\\
k_{\frac{3}{2}}(r) = \alpha & \left(1+\sqrt{\sum_{i=1}^{D}\frac{3(x_i-x_i')^2}{\ell_i^2}}\;\right)\\
& \times \exp\! \left( - \sqrt{\sum_{i=1}^{D}\frac{3(x_i-x_i')^2}{\ell_i^2}}\;\right).
\end{align*}
%
Notice that the previous expressions, $k_{\infty}(r)$ and $k_{\frac{3}{2}}(r)$, have been easily generalized to using a multidimensional length-scale $\bm{\ell}\in {\rm I\!R}^D$. Using individual length-scales for each dimension turns the isotropic covariance function into a non-isotropic covariance function. That is, for a non-isotropic covariance function, the smoothness may vary across different input dimensions.

Stationary covariance functions can be represented in terms of their spectral densities \citep[see, e.g.,][]{rasmussen2006gaussian}. In this sense, the covariance function of a stationary process can be represented as the Fourier transform of a positive finite measure \citep[\textit{Bochner's theorem}; see, e.g., ][]{akhiezer1993theory}. If this measure has a density, it is known as the spectral density of the covariance function, and the covariance function and the spectral density are Fourier duals, known as the \textit{Wiener-Khintchine theorem} \citep{rasmussen2006gaussian}. The spectral density functions associated with the Mat\'ern class of covariance functions are given by
%
\begin{align*}
S_{\nu}(\bm{\omega})&= \alpha \, \frac{2^D\pi^{D/2}\Gamma(\nu+D/2)(2\nu)^{\nu}}{\Gamma(\nu)\, \ell^{2\nu}}\left(\frac{2\nu}{\ell^2}+4\pi^2\bm{\omega}^\intercal \bm{\omega} \right)^{(\nu+D/2)}
\end{align*}
%
in $D$ dimensions, where vector $\bm{\omega}\in {\rm I\!R^D}$ is in the frequency domain, and $\ell$ and $\alpha$ are the length-scale and magnitude (marginal variance), respectively, of the kernel. The particular cases, where $\nu=\infty$ and $\nu=3/2$, take the form
%
\begin{align}
S_{\infty}(\bm{\omega})&= \alpha\,(\sqrt{2\pi})^D  \ell^D \exp(-\frac{1}{2}\ell^2 \bm{\omega}^\intercal \bm{\omega}), \label{eq_specdens_inf}  \\
S_{\frac{3}{2}}(\bm{\omega})&= \alpha\,\frac{2^D\pi^{D/2}\Gamma(\frac{D+3}{2}){3}^{3/2}}{\frac{1}{2}\sqrt{\pi}\ell^3}\left(\frac{3}{\ell^2}+\bm{\omega}^\intercal \bm{\omega} \right)^{-\frac{D+3}{2}}. \label{eq_specdens_32} 
\end{align}
%
For instance, with input dimensionality $D=3$ and $\bm{\omega}=(\omega_1,\omega_2,\omega_3)^\intercal$, the spectral densities written above take the form
%
\begin{align*}
S_{\infty}(\bm{\omega})&= \alpha \, (2\pi)^{3/2}  \prod_{i=1}^{3} \ell_i  \exp\!\left(-\frac{1}{2} \sum_{i=1}^{3} \ell_i^2 \omega_i^2 \right),   \\
S_{\frac{3}{2}}(\bm{\omega})&= \alpha \, 32\,\pi {3}^{3/2}\prod_{i=1}^{3}\ell_i\left(3+\sum_{i=1}^{3}\ell_i^2 \omega_i^2 \right)^{-3},
\end{align*}
%
where individual length-scales $\ell_i$ for each frequency dimension $\omega_i$ have been used.

@@

The accuracy and speed of the HSGP model depends on several interrelated factors, most notably on the number of basis functions and on the boundary condition of the Laplace eigenfunctions. Furthermore, appropriate values for these factors will depend on the degree of non-linearity of the function to be estimated, which is in turn is characterized by the length-scale of the covariance function. In this section, we analyze the effects of the number of basis functions and the boundary condition on the approximation accuracy. We present recommendations on how they should be chosen and diagnostics to check the accuracy of the obtained approximation.

@@

\begin{figure*}[!t]
\centering
\subfigure{\includegraphics[scale=0.29, trim = 1mm 5mm 3mm 0mm, clip]{fig1_Post_J.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 0mm 5mm 3mm 0mm, clip]{fig1_Sigma_J.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 1mm 5mm 3mm 0mm, clip]{fig1_Cov_J.pdf}}
\vspace{-1mm}
\caption{Mean posterior predictive functions (left), posterior standard deviations (center), and covariance functions (right) of both the exact GP model (dashed red line) and the HSGP model for different number of basis functions $m$, with the boundary factor fixed to a large enough value.}
  \label{fig1_Post_J}
\end{figure*}

@@

\begin{figure*}[!t]
\centering
\subfigure{\includegraphics[scale=0.38, trim = 0mm 8mm 0mm 0mm, clip]{fig4_MSE_vs_J.pdf}}
\subfigure{\includegraphics[scale=0.38, trim = 0mm 8mm 0mm 0mm, clip]{fig4_MSE_vs_c.pdf}}
\caption{Root mean square error (RMSE) of the proposed HSGP models computed against the exact GP model. RMSE versus the number of basis functions $m$ and for different values of the boundary factor $c$ (left). RMSE versus the boundary factor $c$ and for different values of the number of basis functions $m$ (right). }
  \label{fig4_MSE_vs_J}
\end{figure*}

\begin{figure*}[!h]
\centering
\subfigure{\includegraphics[scale=0.38, trim = 0mm 8mm 0mm 0mm, clip]{fig5_lscale_vs_J.pdf}}
\subfigure{\includegraphics[scale=0.38, trim = 0mm 8mm 0mm 0mm, clip]{fig5_magnitud_vs_J.pdf}}
\caption{Estimated length-scale (left) and marginal variance (right) parameters of both exact GP and HSGP models, plotted versus the number of basis functions $m$ and for different values of the boundary factor $c$.}
  \label{fig5_lscale_vs_J}
\end{figure*}

Additionally, there is a clear relation between the number of basis functions $m$ and the boundary factor $c$ with the length-scale $\ell$ of the approximated function. Figures \ref{fig6_lscale_vs_J_vs_c} and \ref{fig7_lscale_vs_J_vs_c_Matern} depict how these three factors interact and affect the accuracy of the HSGP approximation for a GP with square exponential covariance function and Mat\'ern ($\nu$=3/2) covariance function, respectively, and a single input dimension. More precisely, for a given GP model (with a squared exponential covariance function) with length-scale $\ell$ and given a boundary factor $c$, Figure \ref{fig6_lscale_vs_J_vs_c} shows the minimum number of basis functions $m$ required to obtain an accurate approximation in the sense of satisfying eq. (\ref{eq_diff_covs}). Similarly,  Figure \ref{fig7_lscale_vs_J_vs_c_Matern} shows the corresponding plot for the  Mat\'ern ($\nu$=3/2) covariance function. We considered an approximation to be a sufficiently accurate when the total variation difference between the approximate and exact covariance functions, $\varepsilon$ in eq. (\ref{eq_diff_covs}), is below 1$\%$ of the total area under the curve of the exact covariance function $k$ such that
%
\begin{equation*}
 \int | k(\tau) - \tilde{k}_m(\tau)|  \,\mathrm{d}\tau  < 0.01 \int k(\tau) \,\mathrm{d}\tau,
\end{equation*}
where $\tilde{k}_m$ is the approximate covariance function with $m$ basis functions. Alternatively, these figures can be understood as providing the minimum $c$ that we should use for given $\ell$ and $m$. Of course, we may also read it as providing the minimum $\ell$ that can be approximated with high accuracy given $m$ and $c$. We obtain the following main conclusions:

\begin{itemize}
\item As $\ell$ increases, $m$ required for an accurate approximation decrease.
\item The lower $c$, the smaller $m$ can and $\ell$ must be to achieve an accurate approximation.
\item For a given $\ell$ there exist a minimum $c$ under which a accurate approximation is never going to be achieved regardless of $m$. This fact can be seen in Figures~\ref{fig6_lscale_vs_J_vs_c} and \ref{fig7_lscale_vs_J_vs_c_Matern} as the contour lines which represent $c$ have an end in function of $\ell$ (Valid $c$ are restricted in function of $\ell$). A $\ell$ increases, the minimum valid $c$ also increases.
\end{itemize}

As stated above, Figures~\ref{fig6_lscale_vs_J_vs_c} and \ref{fig7_lscale_vs_J_vs_c_Matern} provide the minimum length-scale that can be closely approximated given $m$ and $c$. This information serves as a powerful diagnostic tool in determining if the obtained accuracy is acceptable. As the length-scale $\ell$ controls the wigglyness of the function, it strongly influences the difficulty of obtaining accurate inference about the function from the data. Basically, if the length-scale estimate is accurate, we can expect the HSGP approximation to be accurate as well.

Having obtained an estimate $\hat{\ell}$ of $\ell$ from the HSGP model based on prespecified $m$ and $c$, we can check whether or not $\hat{\ell}$ exceeds the minimum length-scale provided in Figure~\ref{fig6_lscale_vs_J_vs_c} or \ref{fig7_lscale_vs_J_vs_c_Matern} (depending on which kernel is used). If $\hat{\ell}$ exceeds this recommended minimum length-scale, the approximation is assumed to be good. If, however, $\hat{\ell}$ does not exceed recommended minimum length-scale, the approximation may be inaccurate and $m$ should be increased or $c$ decreased. We may also use this diagnostic in an iterative procedure by starting from some initial guess of $\ell$ and initial values for $m$ and $c$, and if the estimated $\hat{\ell}$ is below the minimum length-scale, repeat the process while increasing $m$ or decreasing $c$. As mentioned earlier, $c$ cannot be decreased too much as the lowest useful value of $c$ is restricted by the length-scale. Thus, increasing $m$ may usually the preferred approach.

\begin{figure*}
\centering
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 5mm 0mm, clip]{fig6_lscale_vs_J_vs_c.pdf}}
\hspace{3mm}
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 5mm 0mm, clip]{fig6_lscale_vs_J_vs_c_zoomin.pdf}}
\caption{Relation among the minimum number of basis functions $m$, the boundary factor $c$ \hspace{0.1mm} ($c = \frac{L}{S}$) \hspace{0.1mm} and the length-scale normalized by the half-range of the data ($\frac{\ell}{S}$), in the case of a square exponential covariance function. The right-side plot is a zoom in of the left-side plot.}
  \label{fig6_lscale_vs_J_vs_c}
\end{figure*}

\begin{figure*}
\centering
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 5mm 0mm, clip]{fig7_lscale_vs_J_vs_c_Matern.pdf}}
\hspace{3mm}
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 5mm 0mm, clip]{fig7_lscale_vs_J_vs_c_zoomin_Matern.pdf}}
\caption{Relation among the minimum number of basis functions $m$, the boundary factor $c$ \hspace{0.1mm} ($c = \frac{L}{S}$) \hspace{0.1mm} and the length-scale normalized by the half-range of the data ($\frac{\ell}{S}$), in the case of a Mat\'ern ($\nu$=3/2) covariance function. The right-side plot is a zoom in of the left-side plot.}
  \label{fig7_lscale_vs_J_vs_c_Matern}
\end{figure*}

If we look back to the conclusions drawn from Figures~\ref{fig4_MSE_vs_J} and \ref{fig5_lscale_vs_J}, where $m = 10$ basis functions and a boundary factor of $c = 1.5$ were sufficient to obtain an accurate approximation of a function with $\ell = 0.3$, we can recognize that these conclusions also matches those obtained from Figure~\ref{fig6_lscale_vs_J_vs_c}.

>>>>

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
%\begin{filecontents*}{example.eps}
%%!PS-Adobe-3.0 EPSF-3.0
%%%BoundingBox: 19 19 221 221
%%%CreationDate: Mon Sep 29 1997
%%%Creator: programmed by hand (JK)
%%%EndComments
%gsave
%newpath
%  20 20 moveto
%  20 220 lineto
%  220 220 lineto
%  220 20 lineto
%closepath
%2 setlinewidth
%gsave
%  .4 setgray fill
%grestore
%stroke
%grestore
%\end{filecontents*}
%
%\RequirePackage{fix-cm}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
%\documentclass[smallextended]{svjour3}       % onecolumn (second format)
\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
\usepackage{blindtext}
%\usepackage{titlesec}
\usepackage[utf8]{inputenc}
\usepackage{times}
\usepackage{microtype}
\usepackage{lscape}
\usepackage{epstopdf}
\usepackage{natbib}
\usepackage{subfigure}
\usepackage{multirow}
\usepackage{multicol}
\usepackage{float}
\usepackage{soul}
\usepackage{xcolor}
\usepackage{amssymb}
\usepackage{amsmath}	
\usepackage{mathtools}
%\usepackage{amsthm}
\usepackage{bm}
\usepackage{url}
\usepackage{parskip}
\usepackage{pbox}
\usepackage[titletoc]{appendix}
\usepackage{booktabs}
\usepackage{boldline}
\usepackage{colortbl}
\usepackage{listings}
\usepackage{caption}
\usepackage[colorlinks=true,citecolor=black,linkcolor=black,urlcolor=black,breaklinks=true]{hyperref}
\usepackage{nicefrac}

@@

{\color{blue} Aki: we could cite in the intro https://arxiv.org/abs/2111.01084 as a paper which
reviews alternative methods}

@@

The Mat\'ern class of isotropic covariance functions is given by, 
%
\begin{align*}
k_{\nu}(r)&=\alpha \, \frac{2^{1-\nu}}{\Gamma(\nu)}\left(\frac{\sqrt{2\nu}r}{\ell}\right)^{\!\nu} \! K_{\nu} \left(\frac{\sqrt{2\nu}r}{\ell}\right),
\end{align*}
%
where $\nu > 0$ is the order the kernel, $K_{\nu}$ the modified Bessel function of the second kind, and the $\ell > 0$ and $\alpha > 0$ are the length-scale and magnitude (marginal variance), respectively, of the kernel. The particular case where $\nu=\infty$, $\nu=1/2$, $\nu=3/2$ and $\nu=5/2$ are probably the most commonly used kernels \citep{rasmussen2006gaussian}, 
%
\begin{align*}
k_{\infty}(r)&=\alpha \exp \left( -\frac{1}{2} \frac{r^2}{\ell^2}\right),  \\
k_{\frac{1}{2}}(r)&=\alpha \exp \left( -\frac{r}{\ell}\right),  \\
k_{\frac{3}{2}}(r)&=\alpha\left(1+\frac{\sqrt{3}r}{\ell}\right) \exp\! \left( -\frac{\sqrt{3}r}{\ell}\right),   \\
k_{\frac{5}{2}}(r)&=\alpha\left(1+\frac{\sqrt{5}r}{\ell} + \frac{5r^2}{3\ell^2}\right) \exp\! \left( -\frac{\sqrt{5}r}{\ell}\right). 
\end{align*}
%
The former is commonly known as the squared exponential or exponentiated quadratic covariance function. Assuming the Euclidean distance between observations, $r=|\bm{x}-\bm{x}'|_{L2}=\sqrt{\sum_{i=1}^{D}(x_i-x_i')^2}$, the kernels written above take the form
%
\begin{align*}
k_{\infty}(r) = \alpha & \exp \left( -\frac{1}{2} \sum_{i=1}^{D}\frac{(x_i-x_i')^2}{\ell_i^2}\right),\\
k_{\frac{1}{2}}(r) = \alpha & \exp \left( -\sum_{i=1}^{D}\frac{(x_i-x_i')}{\ell_i}\right),\\
k_{\frac{3}{2}}(r) = \alpha & \left(1+\sqrt{\sum_{i=1}^{D}\frac{3(x_i-x_i')^2}{\ell_i^2}}\;\right)\\
& \times \exp\! \left( - \sqrt{\sum_{i=1}^{D}\frac{3(x_i-x_i')^2}{\ell_i^2}}\;\right),\\
k_{\frac{5}{2}}(r) = \alpha & \left(1+\sqrt{\sum_{i=1}^{D}\frac{3(x_i-x_i')^2}{\ell_i^2}}+\frac{5}{3} \sum_{i=1}^{D}\frac{(x_i-x_i')^2}{\ell_i^2}\;\right)\\
& \times \exp\! \left( - \sqrt{\sum_{i=1}^{D}\frac{5(x_i-x_i')^2}{\ell_i^2}}\;\right).
\end{align*}
%
Notice that the previous expressions, $k_{\infty}(r)$, $k_{\frac{1}{2}}(r)$, $k_{\frac{3}{2}}(r)$ and $k_{\frac{5}{2}}(r)$, have been easily generalized to using a multidimensional length-scale $\bm{\ell}\in {\rm I\!R}^D$. Using individual length-scales for each dimension turns the isotropic covariance function into a non-isotropic covariance function. That is, for a non-isotropic covariance function, the smoothness may vary across different input dimensions.

Stationary covariance functions can be represented in terms of their spectral densities \citep[see, e.g.,][]{rasmussen2006gaussian}. In this sense, the covariance function of a stationary process can be represented as the Fourier transform of a positive finite measure \citep[\textit{Bochner's theorem}; see, e.g., ][]{akhiezer1993theory}. If this measure has a density, it is known as the spectral density of the covariance function, and the covariance function and the spectral density are Fourier duals, known as the \textit{Wiener-Khintchine theorem} \citep{rasmussen2006gaussian}. The spectral density functions associated with the Mat\'ern class of covariance functions are given by
%
\begin{align*}
S_{\nu}(\bm{\omega})&= \alpha \, \frac{2^D\pi^{D/2}\Gamma(\nu+D/2)(2\nu)^{\nu}}{\Gamma(\nu)\, \ell^{2\nu}}\left(\frac{2\nu}{\ell^2}+4\pi^2\bm{\omega}^\intercal \bm{\omega} \right)^{-(\nu+D/2)}
\end{align*}
%
in $D$ dimensions, where vector $\bm{\omega}\in {\rm I\!R^D}$ is in the frequency domain, and $\ell$ and $\alpha$ are the length-scale and magnitude (marginal variance), respectively, of the kernel. The particular cases, where $\nu=\infty$, $\nu=1/2$, $\nu=3/2$ and $\nu=5/2$, take the form
%
\begin{align}
S_{\infty}(\bm{\omega})&= \alpha\,(\sqrt{2\pi})^D  \ell^D \exp(-\frac{1}{2}\ell^2 \bm{\omega}^\intercal \bm{\omega}), \label{eq_specdens_inf}  \\
S_{\frac{1}{2}}(\bm{\omega})&= \alpha\,\frac{2^D\pi^{D/2}\Gamma(\frac{D+1}{2})}{\sqrt{\pi}\ell}\left(\frac{1}{\ell^2}+\bm{\omega}^\intercal \bm{\omega} \right)^{-\frac{D+1}{2}} \label{eq_specdens_12},     \\
S_{\frac{3}{2}}(\bm{\omega})&= \alpha\,\frac{2^D\pi^{D/2}\Gamma(\frac{D+3}{2}){3}^{3/2}}{\frac{1}{2}\sqrt{\pi}\ell^3}\left(\frac{3}{\ell^2}+\bm{\omega}^\intercal \bm{\omega} \right)^{-\frac{D+3}{2}} \label{eq_specdens_32},     \\
S_{\frac{5}{2}}(\bm{\omega})&= \alpha\,\frac{2^D\pi^{D/2}\Gamma(\frac{D+5}{2}){5}^{5/2}}{\frac{3}{4}\sqrt{\pi}\ell^5}\left(\frac{5}{\ell^2}+\bm{\omega}^\intercal \bm{\omega} \right)^{-\frac{D+5}{2}}. \label{eq_specdens_52}   
\end{align}
%
For instance, with input dimensionality $D=3$ and $\bm{\omega}=(\omega_1,\omega_2,\omega_3)^\intercal$, the spectral densities written above take the form
%
\begin{align*}
S_{\infty}(\bm{\omega})&= \alpha \, (2\pi)^{3/2}  \prod_{i=1}^{3} \ell_i  \exp\!\left(-\frac{1}{2} \sum_{i=1}^{3} \ell_i^2 \omega_i^2 \right),   \\
S_{\frac{1}{2}}(\bm{\omega})&= \alpha \, 8\,\pi \prod_{i=1}^{3}\ell_i\left(1+\sum_{i=1}^{3}\ell_i^2 \omega_i^2 \right)^{-2},    \\
S_{\frac{3}{2}}(\bm{\omega})&= \alpha \, 32\,\pi {3}^{3/2}\prod_{i=1}^{3}\ell_i\left(3+\sum_{i=1}^{3}\ell_i^2 \omega_i^2 \right)^{-3},   \\
S_{\frac{5}{2}}(\bm{\omega})&= \alpha \, \frac{64}{3}\,\pi {5}^{5/2}\prod_{i=1}^{3}\ell_i\left(5+\sum_{i=1}^{3}\ell_i^2 \omega_i^2 \right)^{-4}.
\end{align*}
%
where individual length-scales $\ell_i$ for each frequency dimension $\omega_i$ have been used.

@@

The accuracy and speed of the HSGP model depends on several interrelated factors, most notably on the number of basis functions and on the boundary condition of the Laplace eigenfunctions. Furthermore, appropriate values for these factors will depend on the degree of non-linearity of the function to be estimated, which is in turn characterized by the length-scale of the covariance function. In this section, we analyze the effects of the number of basis functions and the boundary condition on the approximation accuracy. We present recommendations on how they should be chosen and diagnostics to check the accuracy of the obtained approximation.

@@

\begin{figure*} %[!t]
\centering
\subfigure{\includegraphics[scale=0.29, trim = 1mm 5mm 3mm 0mm, clip]{fig1_Post_J.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 0mm 5mm 3mm 0mm, clip]{fig1_Sigma_J.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 1mm 5mm 3mm 0mm, clip]{fig1_Cov_J.pdf}}
\vspace{-1mm}
\caption{Mean posterior predictive functions (left), posterior standard deviations (center), and covariance functions (right) of both the exact GP model (dashed red line) and the HSGP model for different number of basis functions $m$, with the boundary factor fixed to a large enough value.}
  \label{fig1_Post_J}
\end{figure*}

@@

\begin{figure*} %[!t]
\centering
\subfigure{\includegraphics[scale=0.38, trim = 0mm 8mm 0mm 0mm, clip]{fig4_MSE_vs_J.pdf}}
\subfigure{\includegraphics[scale=0.38, trim = 0mm 8mm 0mm 0mm, clip]{fig4_MSE_vs_c.pdf}}
\caption{Root mean square error (RMSE) of the proposed HSGP models computed against the exact GP model. RMSE versus the number of basis functions $m$ and for different values of the boundary factor $c$ (left). RMSE versus the boundary factor $c$ and for different values of the number of basis functions $m$ (right). }
  \label{fig4_MSE_vs_J}
\end{figure*}

\begin{figure*} %[!h]
\centering
\subfigure{\includegraphics[scale=0.38, trim = 0mm 8mm 0mm 0mm, clip]{fig5_lscale_vs_J.pdf}}
\subfigure{\includegraphics[scale=0.38, trim = 0mm 8mm 0mm 0mm, clip]{fig5_magnitud_vs_J.pdf}}
\caption{Estimated length-scale (left) and marginal variance (right) parameters of both exact GP and HSGP models, plotted versus the number of basis functions $m$ and for different values of the boundary factor $c$.}
  \label{fig5_lscale_vs_J}
\end{figure*}

\subsection{Theoretical evidence of linear proportionality between $m$, $c$ and $l$} \label{subsec_theoretic_linearity}

There is a clear relation between the number of basis functions $m$ and the boundary factor $c$ with the length-scale $\ell$ of the approximated function. In this Section we discuss about the existing theoretical evidence that the minimum number of basis functions $m$ needed for an accurate approximation is linearly proportional to the the boundary factor $c$ and inversely proportional to the lengthscale $l$. In next Section \ref{subsec_empiric_relations} we conduct an empirical experiments to prove this expected linear proportionality relationship between $m$, $c$ and $l$ and compute the specific coefficients of proportionality for every kernel.

{\color{blue}Aki: We can also look at the mean zero-level upcrossings, which for QE is
proportional to 1/l which gives as that part, and the effect of c should also
be quite linear when c is not very close to 1 and m is not very small as it
linearly makes the interval shorter so that the there are less zero-level
upcrossings available in the basis functions. You can visualize this using
{\small \url{https://avehtari.github.io/casestudies/Motorcycle/motorcycle_gpcourse.html}} and
look how the plot of 6 basis functions change when you set c=1, c=1.5, c=2.25
(ie $1.5^2$). Instead of looking at mean zero-level crossings, we can look at the
spectral densities and when they go to zero. This is illustrated in the same
notebook after that figure. E.g. with lengthscale l=1,l=1/2, l=1/4, l=1/8 to
explain 99\% of the prior variance we need 4, 8, 16, and 32 basis functions
(linear in l) and to explain 80\% of the prior variance we need 2, 4, 8, 16
basis functions. These spectral densities don't depend on c. When c increaes
the basis functions are truncated and for example going from c=1.5 to c=2.25,
basis function 9 has the flexibility basis function 6 had, so 1.5 more basis
functions are needed. So we have proportionality c/l. }

{\color{red} Gabi: The spectral densities do depend on $c$, see equations (\ref{eq_approxf}) and (\ref{eq_eigenvalue}).}

\subsection{Empirical discovering of the functional form of the relationships between $m$, $c$ and $l$} \label{subsec_empiric_relations}

Figures \ref{fig6_lscale_vs_J_vs_c} and \ref{fig6_7_relationships} depict how these three factors interact and affect the accuracy of the HSGP approximation for a GP with square exponential covariance function and Mat\'ern ($\nu$=3/2) covariance function, respectively, and a single input dimension. More precisely, for a given GP model (with a squared exponential covariance function) with length-scale $\ell$ and given a boundary factor $c$, Figure \ref{fig6_lscale_vs_J_vs_c} shows the minimum number of basis functions $m$ required to obtain an accurate approximation in the sense of satisfying eq. (\ref{eq_diff_covs}). Similarly,  Figure \ref{fig6_7_relationships} shows the corresponding plot for the  Mat\'ern ($\nu$=3/2) covariance function. We considered an approximation to be a sufficiently accurate when the total variation difference between the approximate and exact covariance functions, $\varepsilon$ in eq. (\ref{eq_diff_covs}), is below 1$\%$ of the total area under the curve of the exact covariance function $k$ such that
%
\begin{equation*}
 \int | k(\tau) - \tilde{k}_m(\tau)|  \,\mathrm{d}\tau  < 0.01 \int k(\tau) \,\mathrm{d}\tau,
\end{equation*}
where $\tilde{k}_m$ is the approximate covariance function with $m$ basis functions. Alternatively, these figures can be understood as providing the minimum $c$ that we should use for given $\ell$ and $m$. Of course, we may also read it as providing the minimum $\ell$ that can be approximated with high accuracy given $m$ and $c$. We obtain the following main conclusions:

\begin{itemize}
\item As $\ell$ increases, $m$ required for an accurate approximation decrease.
\item The lower $c$, the smaller $m$ can and $\ell$ must be to achieve an accurate approximation.
\item For a given $\ell$ there exist a minimum $c$ under which a accurate approximation is never going to be achieved regardless of $m$. This fact can be seen in Figures~\ref{fig6_lscale_vs_J_vs_c} and \ref{fig6_7_relationships} as the contour lines which represent $c$ have an end in function of $\ell$ (Valid $c$ are restricted in function of $\ell$). As $\ell$ increases, the minimum valid $c$ also increases.
\end{itemize}

\subsubsection{Numerical equations}

From the empirical observations the hypothesis of linear proportionality between $m$, $\frac{l}{S}$ (lenghscale $l$ relative to the half range of the input domain $S$) and $c$ has been proved. Following we show numerical functions governing these relationships for every kernel.

\paragraph*{Squared exponential}:
%
\begin{align}\label{eq_m_l_QE}
&m= 1.75 \, \frac{c}{\nicefrac{l}{S}} \;\; \Leftrightarrow \;\; \nicefrac{l}{S}= 1.75 \, \frac{c}{m},
\end{align}
%
with
%
\begin{align}\label{eq_c_vs_l_QE}
&c = c(l) \, \geq \, 3.2 \, \nicefrac{l}{S}         
\end{align}

\paragraph*{Mat\'ern ($\nu$=3/2)}:
%
\begin{align}\label{eq_m_l_mat32}
&m= 3.42 \, \frac{c}{\nicefrac{l}{S}} \;\; \Leftrightarrow \;\; \nicefrac{l}{S}= 3.42 \, \frac{c}{m},
\end{align}
%
with
%
\begin{align}\label{eq_c_vs_l_mat32}
&c \, \geq \, 4.5 \, \nicefrac{l}{S}         
\end{align}

\paragraph*{Mat\'ern ($\nu$=5/2)}:
%
\begin{align}\label{eq_m_l_mat52}
&m= 2.65 \, \frac{c}{\nicefrac{l}{S}} \;\; \Leftrightarrow \;\; \nicefrac{l}{S}= 2.65 \, \frac{c}{m},
\end{align}
%
with
%
\begin{align}\label{eq_c_vs_l_mat52}
&c \, \geq \, 4.1 \, \nicefrac{l}{S}         
\end{align}
